---
title: "Stroke Mortality - ANOVA"
author: "Author: Marissa Munoz-Ruiz"
output: html_document
---

$~$

#### **Goal of Script:** Explore ethnicity in stroke mortality 

Is there is a difference in stroke mortality among different ethnic groups? Does stroke mortality differ in various regions of the U.S.?

The U.S. government has publicly accessible data on stroke mortality on data.gov. I downloaded the excel file for 2016 ([Stroke Mortality](https://chronicdata.cdc.gov/Heart-Disease-Stroke-Prevention/Stroke-Mortality-Data-Among-US-Adults-35-by-State-/v246-z5tb)). Since the stroke dataset gives the stroke mortality rate across ethnicity and gender at the state level, I also downloaded an additional dataset that breaks down states into different regions ([Regions](https://www.kaggle.com/omer2040/usa-states-to-region)). 

I will use ggplot2 to visually explore the data and then conduct an ANOVA analysis to determine if there is a statistical difference between groups. Note: All plots use the same color schemes for consistent representation across ethnic groups and regions. 

```{r setup, message=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)

library(readr)
library(dplyr)
library(ggplot2)
library(rstatix) 

# Load Data (only 2016)
file <- "C:/Users/mmuno/Desktop/GitHub/ores5310-2020 (Stats)/Data"
Data <- read_csv(paste(file,'/StrokeMortality_2016.csv',sep=''))
USRegions <- read_csv(paste(file,'/State_Regions.csv',sep='')) 

# Define colors for plotting
EthnList = c("Asian and Pacific Islander","Black","Hispanic","White")
EthnColor = c("darkorange","purple","red","darkgreen")
RegList = c("Midwest","Northeast","South","West")
RegColor= c("gray30","darkgoldenrod3","deeppink","darkcyan")

List = c("Midwest"="gray30","Northeast"="darkgoldenrod3","South"="deeppink","West"="darkcyan","Asian and Pacific Islander"="darkorange","Black"="purple","Hispanic"="red","White"="darkgreen")

# Trim Data
USRegions <- USRegions %>% 
  select(-State) %>%
  rename(StateCode='State Code') 

StrokeData <- Data %>% 
  filter(GeographicLevel=="State",!is.na(Data_Value)) %>%
  select(LocationAbbr, Data_Value, Stratification1, Stratification2) %>%
  rename(StateCode=LocationAbbr, MortalityRate=Data_Value, Sex=Stratification1, Ethnicity=Stratification2)%>%
  left_join(USRegions,"StateCode")

#Basic statistics
Stroke_Stats <- StrokeData %>%
  select(Region, Ethnicity, MortalityRate) %>% 
  filter(Ethnicity!="Overall", Ethnicity!="American Indian and Alaskan Native") %>%  
  mutate(Region=factor(Region,levels=RegList), Ethnicity=factor(Ethnicity,levels=EthnList)) %>%
  group_by(Region, Ethnicity) %>%
  summarize(MR=round(mean(MortalityRate, na.rm=TRUE),3),
            sd=round(sd(MortalityRate),3),
            skew=round(e1071::skewness(MortalityRate),3),
            N=n()) %>%
  arrange(Region, Ethnicity)

Stroke_Stats

```
At first glance, the data shows that Black ethnic group has the highest mortality rates in the South and in the Midwest. 

Note: The American Indian and Alaskan Native ethnicity group was removed from the data because mortality rate for this ethnic group was not listed at all levels of region. There were two states that were not listed in the Regions file and those were also removed. 

#### **Data Visualization**

```{r}

# Bar graph - Ethnicity & Region
Stroke_Stats %>%
ggplot(aes(x = Region, y=MR, fill=Ethnicity)) +
  geom_col(width=0.65, position=position_dodge(0.8), colour="black") +
  geom_errorbar(aes(ymin=MR-sd, ymax=MR+sd),width=0.2,position=position_dodge(0.8))+
  scale_fill_manual(values=EthnColor) +  
  labs(x="Region", y="Mortality Rate", title="Fig 1: Stroke Mortality among Ethnic Groups & Region") +
  theme(plot.title = element_text(hjust=0.5))

```


```{r}
# Interaction Plot
Stroke_Stats %>% 
  ggplot(aes(x = Ethnicity, y = MR)) +
  geom_line(aes(group=Region,color=Region)) +
  geom_point(aes(color=Ethnicity),size=4) +
  scale_color_manual(values=List,breaks = RegList) +
  labs(y="Mortality Rate", title="Fig 2: Ethnicity vs Region") +
  theme(plot.title = element_text(hjust=0.5))

```

<br>

```{r}
# Interaction Plot
Stroke_Stats %>% 
  ggplot(aes(x = Region, y = MR)) +
  geom_line(aes(group=Ethnicity,color=Ethnicity)) +
  geom_point(aes(color=Region),size=4) +
  scale_color_manual(values=List, breaks=EthnList) +
  labs(y="Mortality Rate", title="Fig 3: Region vs Ethnicity") +
  theme(plot.title = element_text(hjust=0.5))
```

There seem to be a difference in stroke mortality rate among various ethnic groups and regions (see Fig. 1). The interaction plots indicate there is a complex interaction occurring between ethnic groups and region (see Fig. 2 and Fig. 3).  


```{r}

# Trim data for stats 
Stroke_Analysis <- StrokeData %>%
  select(Region, Ethnicity, MortalityRate) %>% 
  filter(Ethnicity!="Overall", Ethnicity!="American Indian and Alaskan Native") %>% 
  rename(MR=MortalityRate) %>%
  arrange(Region, Ethnicity) 

# Ethnicity distribution
Stroke_Analysis %>%
ggplot(aes(x=MR)) + 
    geom_histogram(binwidth=5) +
    labs(x="Mortality Rate", title = "Fig 4: Distribution of Ethnicity") +
    theme(plot.title = element_text(hjust=0.5)) +
    facet_wrap(~Ethnicity)

```

<br>

```{r}
# Region distribution
Stroke_Analysis %>%
ggplot(aes(x=MR)) + 
    geom_histogram(binwidth=5) +
    labs(x="Mortality Rate", title = "Fig 5: Distribution of Region") +
    theme(plot.title = element_text(hjust=0.5)) +
    facet_wrap(~Region) 
```

<br>

```{r}
# Box plot of Ethnicity & Region
Stroke_Analysis %>%
ggplot(aes(x=Region, y=MR, fill=Ethnicity)) + 
    geom_boxplot() +
    labs(y="Mortality Rate", title = "Fig 6: Ethnicity & Region") +
    scale_fill_manual(values=EthnColor) +
    theme(plot.title = element_text(hjust=0.5)) +
    facet_wrap(~Ethnicity)
```

From the box plots (see Fig 6.), it appears there is a higher stroke mortality rate among Black and White observations in the south region than any other ethnic group or region. Additionally, there seems to be outliers, namely Hispanic observations with really high stroke mortality rates in the West region. There also seems to be a lower stroke mortality rate among Asian and Pacific Islander observations in the Northeast regions compared to other groups and regions.   


#### **Hypotheses**

Ethnicity and region are both independent categorical variables and since I am interested in seeing if there are differences between and among groups then a two way ANOVA is appropriate. 

With informal notation, the null hypothesis for the two way ANOVA with interaction tests are:

1) Group means at any level of Ethnicity are all equal: $$ H_0: \mu_{Alaskan} = \mu_{Black} = \mu_{Hispanic} = \mu_{White}$$

2) Group means at any level of Region are all equal: $$ H_0: \mu_{Midwest} = \mu_{Northeast} = \mu_{South} = \mu_{West}$$

3) There is no interaction effect i.e. effect of Ethnicity does not depend on the effect of Region and vice versa. 

The alternative hypothesis for the two way ANOVA with interaction tests are: 
1) Group means at any level of Ethnicity are not equal 
2) Group means at any level of Region are not equal 
3) There is an interaction effect between Ethnicity and Region

```{r}

# Two Way ANOVA Model
Model <- lm(MR ~ Region*Ethnicity + Region + Ethnicity, data=Stroke_Analysis)
anova(Model)

# Assumptions
par(mfrow=c(2,2))
plot(Model)
shapiro.test(residuals(Model))

Stroke_Analysis %>%
  levene_test(MR ~ Region*Ethnicity) #rstatix function

# Outliers
Stroke_Analysis %>%
  group_by(Region, Ethnicity) %>%
  identify_outliers(MR) 

```

#### **Model Decision**
Since the p-value of the interaction and each factor are all less than our set alpha level of 0.05, we reject all of the null hypotheses. 

#### **Interpretation**

A two-way ANOVA test indicated that the effect of Ethnicity on stroke mortality rate is dependent on the levels of Region and vice versa (F(9,476) = 14.49, p<0.05). Furthermore, there was a main effect for Ethnicity (F(3,476) = 301.03, p<0.05) and Region (F(3,476) = 47.01, p<0.05). 

#### **Assumptions**

In the QQ plot, the residuals look fairly normal. However, the p-value in the Shapiro-Wilk test of normality for the residuals of the model (p=6.47e-12) is significant so normality can't be assumed. In the Residuals vs Fitted plot, there is no relationship between residuals and fitted values so we can assume the homogeneity of variances. The p-value in the Levene Test is significant (p=3.5e-07), which means we can not assume the homogeneity of variances between Ethnicity and Region. There are three potential outliers in the top right hand corner of the QQ plot.   

#### **Pairwise Comparisons** 

```{r}
# Pairwise comparisons
PWC <- Stroke_Analysis %>%
  group_by(Region) %>%
  emmeans_test(MR ~ Ethnicity, p.adjust.method = "bonferroni") #rstatix function

PWC

```

There was a significant difference of stroke mortality rate between all ethnic groups for all levels of Region (p < 0.05). 

